---
title: "Introduction to cori.data.bfs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to cori.data.bfs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval    = TRUE,
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6,
  out.width  = "100%"
)
has_pep <- requireNamespace("cori.data.pep", quietly = TRUE)
```

`cori.data.bfs` provides county, state, and national business application counts
from the U.S. Census Bureau's Business Formation Statistics (BFS), processed and
stored in CORI's S3 data lake. Data cover 2005–present at annual frequency.

## Variables

```{r codebook}
library(cori.data.bfs)

get_bfs_codebook() |> knitr::kable()
```

## Reading data

All data are returned in long format: one row per `geoid / year / variable`.

```{r read-all}
df <- read_bfs_from_s3()
dplyr::glimpse(df)
```

Filter by geography, year, or variable:

```{r read-filtered}
# County-level only, 2010 onward
county <- read_bfs_from_s3(
  years  = 2010:2024
) |>
  dplyr::filter(nchar(geoid) == 5)

dplyr::glimpse(county)
```

## Business applications — rural vs. non-rural

Joining BFS with population from `cori.data.pep` produces per-capita rates.
Joining with `ruraldefinitions` lets us compare rural and non-rural trends.

```{r rural-trend, eval=has_pep, fig.alt="Line chart of business applications per 1000 population, rural vs nonrural, 2005-2024"}
library(ggplot2)
library(cori.charts)
library(dplyr)
library(ruraldefinitions)

load_fonts()

rural_xwalk <- ruraldefinitions::cbsa_2023 |>
  select(geoid, is_rural)

# BFS county totals by rural status
bfs_rural <- read_bfs_from_s3(years = 2005:2024) |>
  filter(nchar(geoid) == 5) |>
  left_join(rural_xwalk, by = "geoid") |>
  filter(!is.na(is_rural)) |>
  group_by(year, is_rural) |>
  summarise(applications = sum(value, na.rm = TRUE), .groups = "drop")

# Population from cori.data.pep
pop_rural <- cori.data.pep::read_pep_from_s3(
  variables = "population",
  years     = 2005:2024
) |>
  filter(nchar(geoid) == 5) |>
  left_join(rural_xwalk, by = "geoid") |>
  filter(!is.na(is_rural)) |>
  group_by(year, is_rural) |>
  summarise(population = sum(value, na.rm = TRUE), .groups = "drop")

chrt_dta <- bfs_rural |>
  left_join(pop_rural, by = c("year", "is_rural")) |>
  mutate(apps_per_1k = applications / population * 1000)

label_df <- chrt_dta |> filter(year == max(year))

ggplot(chrt_dta, aes(x = year, y = apps_per_1k, color = is_rural)) +
  annotate("rect", xmin = 2007.9, xmax = 2009.5,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.4) +
  annotate("rect", xmin = 2020, xmax = 2020.5,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.4) +
  geom_line(linewidth = 1.2) +
  ggrepel::geom_text_repel(
    data        = label_df,
    aes(label   = paste0(is_rural, "\n(", round(apps_per_1k, 1), ")")),
    nudge_x     = 0.5,
    direction   = "y",
    hjust       = 0,
    segment.color = NA,
    size        = 3.5,
    lineheight  = 0.9
  ) +
  scale_color_manual(values = c(
    "Rural"    = cori_colors[["Dark Green"]],
    "Nonrural" = cori_colors[["CORI Gray"]]
  )) +
  scale_x_continuous(
    breaks = seq(2005, 2024, by = 4),
    expand = expansion(mult = c(0, 0.2))
  ) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
  theme_cori() +
  theme(legend.position = "none") +
  labs(
    title    = "Rural business formation surged during the pandemic and has held",
    subtitle = "Business applications per 1,000 population, 2005 - 2024",
    x        = NULL,
    y        = NULL,
    caption  = paste0(
      "Source: CORI analysis of U.S. Census Bureau Business Formation Statistics (BFS) ",
      "and Population Estimates \nProgram (PEP).\n",
      "Shaded bands mark Great Recession (2008 - 09) and COVID-19 (2020). ",
      "Rural classification: CORI CBSA 2023 \ndefinition."
    )
  )
```

## Population growth vs. business formation — county scatter

Do counties that are gaining population also see more new business activity?
This chart plots each county's 10-year percent change in population (x-axis)
against its 10-year percent change in business applications (y-axis), colored
by rural status with a trend line for each group.

```{r scatter, eval=has_pep, fig.alt="Scatterplot of population change vs business application change 2014-2024, rural vs nonrural counties"}
# BFS: % change 2014 -> 2024
bfs_change <- read_bfs_from_s3(years = c(2014, 2024)) |>
  filter(nchar(geoid) == 5) |>
  tidyr::pivot_wider(names_from = year, values_from = value, names_prefix = "bfs_") |>
  mutate(bfs_pct_chg = (bfs_2024 - bfs_2014) / bfs_2014 * 100) |>
  filter(!is.na(bfs_pct_chg), is.finite(bfs_pct_chg)) |>
  select(geoid, bfs_pct_chg)

# PEP: % change 2014 -> 2024
pop_change <- cori.data.pep::read_pep_from_s3(
  variables = "population",
  years     = c(2014, 2024)
) |>
  filter(nchar(geoid) == 5) |>
  tidyr::pivot_wider(names_from = year, values_from = value, names_prefix = "pop_") |>
  mutate(pop_pct_chg = (pop_2024 - pop_2014) / pop_2014 * 100) |>
  filter(!is.na(pop_pct_chg), is.finite(pop_pct_chg)) |>
  select(geoid, pop_pct_chg)

scatter_dta <- bfs_change |>
  inner_join(pop_change, by = "geoid") |>
  left_join(rural_xwalk, by = "geoid") |>
  filter(!is.na(is_rural)) |>
  filter(
    bfs_pct_chg >= quantile(bfs_pct_chg, 0.02),
    bfs_pct_chg <= quantile(bfs_pct_chg, 0.98),
    pop_pct_chg >= quantile(pop_pct_chg, 0.02),
    pop_pct_chg <= quantile(pop_pct_chg, 0.98)
  )

ggplot(scatter_dta, aes(x = pop_pct_chg, y = bfs_pct_chg, color = is_rural)) +
  geom_point(alpha = 0.3, size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.4, linetype = "dashed") +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.4, linetype = "dashed") +
  scale_color_manual(
    values = c(
      "Rural"    = cori_colors[["Emerald"]],
      "Nonrural" = cori_colors[["Dark Purple"]]
    ),
    name = NULL
  ) +
  scale_x_continuous(labels = scales::label_percent(scale = 1)) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  theme_cori() +
  theme(legend.position = "bottom") +
  labs(
    title    = "Population growth and business formation move together",
    subtitle = "Business application change, 2014 - 2024",
    x        = "Population change, 2014 - 2024",
    y        = NULL,
    caption  = paste0(
      "Source: CORI analysis of U.S. Census Bureau Business Formation Statistics (BFS) ",
      "and Population Estimates\n Program (PEP).\n",
      "Each point = one county. Trend lines fit by OLS. Extreme outliers (2nd - 98th percentile) trimmed.\nRural classification: CORI CBSA 2023 definition."
    )
  )
```

## County spotlight — Grafton County, NH

```{r grafton, eval=has_pep, fig.alt="Bar chart of business applications in Grafton County NH 2005-2024"}
grafton_bfs <- read_bfs_from_s3(
  geoids = "33009",
  years  = 2005:2024
)

grafton_pop <- cori.data.pep::read_pep_from_s3(
  variables = "population",
  geoids    = "33009",
  years     = 2005:2024
) |>
  select(year, population = value)

grafton <- grafton_bfs |>
  left_join(grafton_pop, by = "year") |>
  mutate(apps_per_1k = value / population * 1000)

ggplot(grafton, aes(x = year, y = apps_per_1k)) +
  annotate("rect", xmin = 2007.9, xmax = 2009.5,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.4) +
  annotate("rect", xmin = 2020, xmax = 2020.5,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.4) +
  geom_col(fill = cori_colors[["Dark Green"]], width = 0.7) +
  scale_x_continuous(breaks = seq(2005, 2024, by = 4)) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
  theme_cori() +
  labs(
    title    = "Business applications per 1,000 residents in Grafton County, NH",
    subtitle = "Annual business applications (BA series), 2005\u20132024",
    x        = NULL,
    y        = NULL,
    caption  = paste0(
      "Source: CORI analysis of U.S. Census Bureau Business Formation Statistics (BFS) ",
      "and Population Estimates \nProgram (PEP).\n",
      "Shaded bands mark Great Recession (2008\u201309) and COVID-19 (2020)."
    )
  )
```
