---
title: "Introduction to cori.data.bfs"
vignette: >
  %\VignetteIndexEntry{Introduction to cori.data.bfs}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval       = FALSE,
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 9,
  fig.height = 6,
  out.width  = "100%"
)
```

`cori.data.bfs` provides annual business application counts from the U.S. Census
Bureau's Business Formation Statistics (BFS). Data cover 2005--present and measure total EIN applications filed with the IRS, excluding tax liens, estates, trusts, agriculture (NAICS 11), and public administration (NAICS 92).

**Source:** U.S. Census Bureau, Business Formation Statistics
**Coverage:** 2005--present, updated annually
**Geography:** County (5-digit FIPS), state (2-digit FIPS), national (`"00"`)

## Variables

```{r codebook, eval=TRUE}
library(cori.data.bfs)
library(dplyr)
library(gt)

get_bfs_codebook() |>
  select(variable, label, unit, notes) |>
  gt() |>
  cols_label(
    variable = "Variable",
    label    = "Label",
    unit     = "Unit",
    notes    = "Notes"
  )


```

## Reading data

All data are returned in long format: one row per `geoid / year / variable`.

```{r read}
df <- get_business_applications(geography = "county")

glimpse(df)
```

Filter to specific geographies, years, or variables:

```{r read-filtered}
# New Hampshire counties, 2010 onward
nh_bfs <- get_business_applications(
  geoids = grep("^33", unique(df$geoid), value = TRUE),
  years  = 2010:2024
)

glimpse(nh_bfs)
```

## Rural vs. Nonrural

Business formation diverges between rural and nonrural counties. The chart below
shows total business applications by rural status using the CBSA 2023 rural
definition.

```{r rural-nonrural}
library(cori.charts)
library(ggplot2)
library(ruraldefinitions)
library(dplyr)

load_fonts()

rural_definitions <- ruraldefinitions::cbsa_2023 |>
  select(geoid, is_rural)

rural_avg <- df |>
  left_join(rural_definitions, by = "geoid") |>
  filter(!is.na(is_rural)) |>
  group_by(year, is_rural) |>
  summarise(value = sum(value, na.rm = TRUE), .groups = "drop")

ggplot(rural_avg, aes(x = year, y = value, color = is_rural)) +
  annotate("rect", xmin = 2007.9, xmax = 2009.5,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.4) +
  annotate("rect", xmin = 2020, xmax = 2020.5,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.4) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_manual(
    values = c("Rural" = "#2F6E9B", "Nonrural" = "#7EBDC2"),
    labels = c("Rural" = "Rural counties", "Nonrural" = "Nonrural counties")
  ) +
  scale_x_continuous(
    breaks = seq(2005, 2024, by = 4),
    expand = expansion(mult = c(0.01, 0.05))
  ) +
  scale_y_continuous(
    labels = scales::label_number(scale_cut = scales::cut_short_scale())
    ) +
  theme_cori() +
  theme(legend.position = "bottom") +
  labs(
    title    = "Rural business formation surged during the pandemic\nand has continued upward",
    subtitle = "Total business applications (BA series), 2005\u20132024",
    x        = NULL,
    y        = NULL,
    color    = NULL,
    caption  = paste0(
      "Source: CORI analysis of U.S. Census Bureau Business Formation Statistics.\n",
      "Shaded bands mark Great Recession (2008\u201309) and COVID-19 (2020).\n",
      "Rural classification: CORI CBSA 2023 definition."
    )
  )
```

## County Spotlight: Grafton County, NH

```{r grafton}
library(cori.data.pep)

grafton_bfs <- get_business_applications(
  geoids = c("33009", "33", "00"),
  years  = 2005:2024
)

grafton_pop <- get_population(
  geoids    = c("33009", "33", "00"),
  variables = "population",
  years     = 2005:2024
) |>
  select(geoid, year, population = value)

grafton <- grafton_bfs |>
  left_join(grafton_pop, by = c("geoid", "year")) |>
  mutate(
    apps_per_1k = value / population * 1000,
    group = case_when(
      geoid == "33009" ~ "Grafton County, NH",
      geoid == "33"    ~ "New Hampshire",
      geoid == "00"    ~ "United States"
    ),
    group = factor(group, levels = c("Grafton County, NH", "New Hampshire", "United States"))
  )

ggplot(grafton, aes(x = year, y = apps_per_1k, color = group, linetype = group, linewidth = group)) +
  annotate("rect", xmin = 2007.9, xmax = 2009.5,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.4) +
  annotate("rect", xmin = 2020, xmax = 2020.5,
           ymin = -Inf, ymax = Inf, fill = "grey80", alpha = 0.4) +
  geom_line(na.rm = TRUE) +
  geom_point(size = 2.5, na.rm = TRUE) +
  scale_color_manual(values = c(
    "Grafton County, NH" = "#2F6E9B",
    "New Hampshire"      = "#7EBDC2",
    "United States"      = "#9DA7B0"
  )) +
  scale_linetype_manual(values = c(
    "Grafton County, NH" = "solid",
    "New Hampshire"      = "dashed",
    "United States"      = "dotted"
  )) +
  scale_linewidth_manual(values = c(
    "Grafton County, NH" = 1.2,
    "New Hampshire"      = 0.8,
    "United States"      = 0.8
  )) +
  scale_x_continuous(breaks = seq(2005, 2024, by = 4)) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1)) +
  theme_cori() +
  theme(legend.position = "bottom") +
  labs(
    title    = "Business applications per 1,000 residents \u2014 Grafton County, NH",
    subtitle = "Compared to New Hampshire and the United States, 2005\u20132024",
    x        = NULL,
    y        = NULL,
    color    = NULL,
    linetype = NULL,
    linewidth = NULL,
    caption  = paste0(
      "Source: CORI analysis of U.S. Census Bureau Business Formation Statistics.\n",
      "Population denominator from Census Population Estimates Program via cori.data.pep.\n",
      "Shaded bands mark Great Recession (2008\u201309) and COVID-19 (2020)."
    )
  )
```
